Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-30 text/back_end_KS.qmd
methods revisions

```{=tex}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}
```
# Introduction

On January 9, 2019 a public research university purchased College Board *Student Search Service* order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, email, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased: first, the 2020 high school graduating class; second, under the heading "College Board Exams," PSAT score range of 1070 - 1180; and third, under the heading "Geography", the order filtered on 67 "Geomarkets," identified by titles like "IL08 - Northwest Suburbs" and "IL09 - North Shore."[^1] The Geomarket filter carves states and metroplitan areas into smaller geographic units. Upon closer inspection, this student list purchase selected all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago," which is significantly poorer and less White.

[^1]: On the same day, the university purchased two additional orders that filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$, respectively, but selected the same HS graduating class and Geomarkets.

```{=html}
<!--Geomarkets carve states and metroplitan areas into smaller geographic units and have names like,  On the same day, the university purchased two additional orders that were identical, except they filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$. 
On the same day, the university also purchased orders $\#448427$ ($57,537$ names) and $\#448440$ ($78,000$ names), which filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$, respectively, but selected the same HS graduating class and Geomarkets as order $\#448922$.

Geomarkets carve states and metroplitan areas into smaller geographic units. Upon closer inspection, the three orders selected on all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago", which is significantly poorer and less White. In the DC/Maryland/Virginia metro area, the orders selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets, which are poorer and have a higher share of Black residents. An analysis of 500 \[check\] student lists purchased by 9 public research universities found that 11% of purchases filtered on Geomarket [@RN5012]. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. To the extent that Geomarket borders are correlated with race and class, the Geomarket selections observed here raise concern that Geomarkets -- and other geographic search filters -- may systematically filter working class and non-white students from educational opportunities.
-->
```
Sociology offers several perspectives to describe the process of sorting students into colleges. The status attainment model argues that student postsecondary destination is a function of family socioeconomic background, particularly parental education [e.g., @RN915; @RN1273; @RN1268; @RN5033; @RN1300]. The cultural capital model argues that bourgeois households bestow the pedigree, social networks, and information necessary for their progeny to claim spots at selective colleges [@RN1228; @RN1241; @RN4991] [ADD CITES]. @RN4873 describes schools as "sorting machines" that create, incorporate, and assign students to categories that determine educational opportunities, often reinforcing reinforcing racial and class inequality. Finally, scholarship on enrollment management offers an agentic depiction of organizational behavior, showing that colleges expend considerable resources recruiting desirable prospects [@RN3519; @RN4758; @cottom2017lower; @RN4324]. All of these perspectives share the view that transitions from school to college tend towards social reproduction. We argue that these perspectives have ignored an important mechanism of social reproduction: third-party vendors create products that sort students on behalf of college, often utilizing historical snapshots of student demand to make recommendations about where colleges should prioritize recruiting efforts. While third-party recruiting products have become increasingly powerful, granular, and ubiquitous in the era of platform capitalism, we argue that a watershed moment in their develeopment was the creation of Geomarkets by the College Board and UPenn professor, Robert Zemsky.

In 1983, Robert Zemsky and Penny Odell authored *The Structure of College Choice*. College Board published the book, provided data, and financed the project, as part of their "efforts to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest geographic unit used in these analyses was the state" [@RN4982, p. x]. @RN4982 developed Geomarkets and the Market Segment Model as an effort to "capture and quantify" (p. 11) the knowledge of admissions officers, based on the idea that "a good recruiter knows where to look for prospective applicants, as seen in the students' willingness or reluctance to travel" (p. 11). Created by analyzing the SAT score-sending behavior of 1980 high school seniors, the thesis of the Market Segment Model is that student demand for your institution is a function of social class and geography. Therefore, colleges should recruit from territories that contain large populations of your target social class. Geomarkets are geographic borders meant to define local recruiting markets, a territory of an admissions recruiter. Figure X, panel A shows Geomarkets in the Chicago-land area. The Market Segment Model predicts how demand for a particular college varies by Geomarket, based on characteristics of households in the Geomarket.

Geomarkets have been incorporated into the supply-side structure of college access in several ways. Albeit anecdotally, Geomarkets became an organizing principle for how college admissions offices allocate admissions recruiters to territories. Figure X panel B shows that Chicago-land Geomarkets define recruiting territories for University of Chicago admissions recruiters (Appendix Figures BLAH BLAH). Second, Geomarkets are the basis for the College Board Enrollment Planning Service (EPS). Founded in 1984 and still active today, EPS software recommends which Geomarkets a college should recruit from and which schools/communities they should prioritize within targeted Geomarkets. According to @RN5021, in 1995, 37% of 4-year public institutions and 49% of 4-year private institutions used EPS, while 41% of 4-year publics and 16% of 4-year privates used ACT's market analysis product, which was based on EPS.

Third, Geomarkets were incorporated into College Board's student list product, Student Search Service. Student lists contain the contact information of prospective students and have been the primary source of lead generation in U.S. higher education since 1972, when College Board began selling names [@belkin2019-studata; @list_biz]. @RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50 in 2021). Colleges control which prospect profiles they purchase by selecting search filters, such as high school graduation year, SAT score, AP score, state, etc. Geomarket filters enable colleges to include/exclude prospects from particular Geomarkets.

Scholarship at the nexus of college access and recruiting has focused on the behavior of colleges [e.g., @cottom2017lower; @RN4758; @RN4759; @RN3519], contributing to the belief that recruiting is a function of individual colleges. Scholarship has failed to investigate the broader enrollment management industry -- consisting of third-party vendors, their products and consulting solutions -- as a set of mechanisms that structure college access. These mechanisms, which lie upstream of the behavior of individual colleges, are incorporated by individual colleges in ways that funnel certain kinds of students to certain kinds of institutions. This omission is surprising because sociology is concerned with sorting [@RN4873], particularly sorting into college.

This manuscript analyzes College Board Geomarkets as a case study of quantification. In particular, we draw from the discussions of correlation and homophily from @RN4975. @RN4982 identified the correlates of 1980 SAT score-sending behavior and concluded that student demand for higher education is primarily a function of social class. The Market Segment Model argues that homophily -- actors that share characteristics form connections -- is the organizing principle of competition and college choice, stating that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students" [@RN4982, p. 72]. Scholarship on quantification demonstrates that making recommendations based on past correlations amplifies the effects of historic structural inequality [@RN4794]. The snapshot of student demand in 1980 -- itself a consequence historic structural inequality -- was programmed into recruiting products that colleges utilize to identify and target prospective students. The result is a supply-side that amplifies structural inequalities observed on the demand-side.

Our analyses address the two research questions, which speak to how Geomarkets are utilized within EPS and within the Student Search Service student list product. First, what is the socioeconomic and racial variation between geomarkets in metropolitan areas and how does this variation change over time? We address this question by spatially joining the geomarket shapefile to Census data about socioeconomic and racial characteristics. Second, how does the socioeconomic and racial composition of included versus included prospects vary when student list purchases filter on particular geomarkets? We address this question by analyzing actual student list purchases that utilize commonly used search filters (e.g., PSAT score, GPA) but do not filter on Geomarkets. We simulate which prospects in particular metropolitan areas would have been included/excluded had the student list purchase filtered on particular Geomarkets. We obtained these data by issuing public records requests to public universities.

The following section provides background about enrollment management and introduces salient concepts from scholarship on quantification. Second, we introduce the case, describing how @RN4982 developed Geomarkets and the Market Segment Model and how these concepts were incorporated into College Board recruiting products. Third, we describe data and methods. Fourth, we present results. [We find that\ldots ?PARAGRAPH ON RESULTS?] Finally, we discuss implications for scholarship and for policy. The sociology of education should move beyond the fixation on schools and policies as the primary sorting machines. @RN527 [p. 148] remind us that the organizational field includes the totality of relevant actors." Third-party vendors have structured college access for decades and their influence is growing. The contemporary enrollment management industry is characterized by private equity owned firms selling software-as-service platforms that utilize the same logic as @RN4982, for example EAB's "pipeline analytics" machine learning product \[CITE\]. By investigating high-leverage actors and products in the organizational field, sociology can develop important insights that shape future federal and state regulatory policy.

```{=html}
\[ADD TEXT SOMEWHERE AROUND HERE ABOUT WHY THIS TOPIC IS SIGNIFICANT AND IMPORTANT FOR CURRENT/FUTURE POLICY DEBATES: Why market segment model important. Software as service products dominate em industry and structure college access. Look at eab machine learning product But rarely do we see the recipe underneath these products. The book structuring college access is that recipe for EPS software\]
<!-- 
THIS TEXT IS FINE, BUT TRYING TO SET UP THE CONTRIBUTION TO BROADER QUANTIFICATION LITERATURE, VIS-A-VIS NO-ONE HAS STUDIED MARKET RESEARCH; DON'T THINK THIS IS NECESSARY AND YOU ARE NOT ON FIRM GROUND HERE IN CALLING THIS "MARKET RESEARCH"
Scholarship on quantification in education has focused on consumer-facing measures such as U.S. News Rankings [@RN5003] and UK "league tables" about school quality [@RN5005]. To our knowledge, prior knowledge has not examined business-facing measures produced by third-party market research. Here, the desirable attributes of students and places they live are quantified with the purpose of informing colleges about where they should allocate recruiting resources to efficiently enroll the most desirable students. We reason that market research reproduces inequality by encouraging colleges to plan their recruiting efforts based on a snapshot of social stratification in student demand that is substantially a function of historic, structural inequality in educational opportunity. In contrast to @RN5005, we cannot demonstrate the effect of geomarkets on inequality. However, we show the mechanisms of how geomarkets reproduce inequality by simulating who is included/excluded when geomarkets are utilized within student list purchases as recommended by @RN4982.
-->
```
# Enrollment Management and Quantification

**Enrollment Management**

<!-- la la -->

Enrollment management is simultaneously a profession, an administrative structure, and an industry. As a profession, enrollment management (EM) integrates techniques from marketing and economics in order to "influence the characteristics and the size of enrolled student bodies" [@RN2771, p. xiv]. As an administrative structure, the EM office typically controls the activities of admissions, financial aid, and recruiting [@RN2406]. <!--The enrollment management industry incorporates relevant actors in the organizational field, including professional associations (e.g., National Association for College Admission Counseling), and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges.  -->

@fig-em-funnel depicts the "enrollment funnel," which modifies the traditional "marketing funnel" to depict broad stages in the process of recruiting students [@RN4986; @RN2004]. The funnel begins with a large pool of "prospects" (i.e., prospective students) that the university would like to enroll. "Leads" are prospects whose contact information has been obtained. "Inquiries" are prospects that contact the institution, including those that respond to an initial solicitation (e.g., email) and those that reach out on their own (e.g., sending SAT scores). <!--  Applicants consist of inquiries who apply plus "stealth applicants" who do not contact the university before applying. The funnel narrows at each successive stage in order to convey the assumption of "melt" (e.g., a subset of "inquiries" will apply).  -->The purpose of the enrollment funnel is to inform recruiting interventions that target one or more stages. These interventions seek to increase the probability of "conversion" across stages [@RN4322]. At the top of the enrollment funnel, purchasing student lists is the primary means of converting prospects to leads [@list_biz]. Purchased leads are served emails, brochures, and targeted social media designed to solicit inquiries and applications [@RN4895].

Scholarship at the nexus of enrollment management and college access can be categorized by which part(s) of the enrollment funnel it speaks to. The majority of scholarship focuses on the admissions stage, analyzing which admissions criteria are utilized and/or which applicants are admitted [e.g., @RN4874; @RN4993; @RN4131; @RN3522; @poon2022rethinking]. Economists often investigate financial aid leveraging, which seeks to convert admits to enrolled students \[CITE\]. <!--  Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935; @RN1814].   -->

A growing literature analyzes the earlier "recruiting" stages of identifying leads, soliciting inquiries, and soliciting applications. @RN4758 conceptualize recruiting behavior as an indicator of college enrollment priorities. Ethnographies by @RN3519 and @RN4407 identify connections between private school guidance counselors and college admissions officers as a mechanism for social reproduction. Recruiting visits to high schools are a means of maintaining ties with guidance counselors at feeder schools and establishing relationships with prospective students [@RN3519; @RN4895]. Quantitative case-studies of off-campus recruiting visits by public research universities and by selective private universities reveal a preference for visiting private schools and affluent, predominantly white public schools [@RN4758; @RN4759; @RN5023]. From the student perspective, @RN4324 finds that underrepresented students were drawn to colleges that made them feel wanted, often attending institutions with lower graduation rates and requiring larger loans than other college options. @cottom2017lower shows that for-profit colleges found a niche in Black and Latinx communities because traditional colleges ignored these communities. <!--  Consistent with scholarship on political economy [@RN4517; @cottom2017lower; @RN5015], this literature reveals a vertical hierarchy of postsecondary institutions matching to a vertical hierarchy of customer niches segmented by class and race, a process consistent with the notion of that higher education is mechanism of social reproduction [@RN2377; @RN4873].   -->

By focusing on the behaviors of individual colleges, scholarship on EM implicitly states that recruiting is something done by individual colleges. In addition to university personnel (e.g., admissions counselors, VP for enrollment management), the EM industry incorporates external stakeholders in the organizational field, including professional associations (e.g., National Association for College Admission Counseling) and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges. We argue college access is structured by third-party servicers and products that interact with direct-providers (colleges). Although sociologists have hinted at the ways enrollment management contributes to inequality in college access [e.g., @RN2406], scholarship has failed to make third-party servicers and products the object of empirical analysis. <!--The enrollment management industry incorporates relevant actors in the organizational field, including professional associations (e.g., National Association for College Admission Counseling), and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges.  --> <!--  
[^1]: New institutional theory offers one explanation [@RN527]. In highly developed organizational fields like higher education, forces of competitive and institutional isomorphism -- including institutional locics -- compel organizations adopt similar practices as their peers. Practices deemed advantageous spread quickly, for example, public research universities targeting nonresident prospects to compensate for declines in state appropriations. These behaviors are conceived as autonomous decisions by individual colleges, albeit structured by institutional logics.
 -->

Drawing from scholarship on organizational theory, enrollment management processes involve many "make or buy" [@RN4738] decisions about whether to perform a given task in-house or outsource it to a third-party vendor [@list_biz]. EM consulting firms provide advice and implementation in the areas of marketing, recruiting, pricing and financial aid, and student success. As "creating a class" has becomes complicated and high-stakes [@RN3519], many colleges hire EM consulting firms to develop and/or implement recruiting campaigns. The two largest firms -- Ruffalo Noel Levitz and EAB -- claim to serve more than 3,000 colleges and universities collectively [@rnlabout; @eabcomm]. In our data collection, attempting to obtain data about student list purchases from all public universities in four states, at least half of these universities outsourced student list purchases to an EM consultancy [@list_biz; @list_empirics].

Although the enrollment management industry is increasingly characterized by software-as-service products sold by private equity backed firms (e.g., for example, EAB's [Enroll 360](https://eab.com/solutions/enroll360/) product) [@list_biz], College Board products played a pivotal role in transforming recruiting from an in-house process to a process structured by third-party products. The Student Search Service, created in 1972, became the ubiquitous means of identifying prospects BLAH BLAH. Second, EPS, launcched in 1984, is an early software-as-service product that provides recommendations about which Geomarkets colleges should recruit from and which high schools they should visit within targeted Geomarkets. Both products are applications of quantification. Student Search Service includes or excludes prospects based on selected attributes while EPS encourages colleges to target Geomarkets and high schools based on the characteristics of households. THEREFORE WE REVIEW SCHOLARSHIP ON QUANTIFICATION.

```{=html}
<!--

The EM industry puts great stock in market research, in part because the EM profession emerged from marketing and market research [@RN4986; @RN2771]. @RN4986 [p. 17] define market research as all "activities intended to increase understanding of a market and consumer behavior in that market." At the most basic level, in-house market research is a function of being intentional about record-keeping and using those records [@RN3519].[^1] In recruiting, market research informs decisions about which prospective students to target with recruiting interventions. How to acquire market research is another make-or-buy decision. The College Board played a pivotal role in transforming market research about enrollment from something done by individual universities to something done by third-party professionals. In 1983, College Board published *Applying Market Research in College Admissions* [@RN4986], which became the basis for their Admitted Student Questionnaire (ASQ) product, and published *The Structure of College Choice* [@RN4982], which became the basis for the Enrollment Planning Service (EPS) product/division. Today, EPS remains the market research arm of College Board, while Encoura Eduventures is the market research arm of ACT. Market research often distills large amounts of data into simpler indicators. Therefore, the quantification literature provides a useful lens to understand market research.

[^1]: @RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."
-->
```
**Quantification**

@RN5007 [p. 402] define quantification as "the production and communication of numbers" (p. 402). Reacivity, discipline, and authority are three interrelated themes that describe the effects of quantification. *Reactivity* is the idea that salient quantitative measures cause people and organizations to change their behavior. Quantification *disciplines* actors to react in particular ways. Quantification changes decision-making power *authority*, often weakening the discretion of local decision-makers. Scholarship about U.S. News & World Report (USNWR) Law School rankings demonstrate the effects of quantification [@RN5003; @RN2356; @RN2376; @RN2375]. Law school rankings were developed as a means of informing prospective students about the relative quality of different programs. Rankings affect (reactivity) where students apply and enroll \[CITE\]. However, rankings also affected hiring decisions of firms because clients evaluated firms based on the prestige of law schools attended by firm lawyers. Once law schools realized that "important groups of constituents --- students, faculty, trustees, employers, other media --- were using rankings to make decisions that had large consequences for schools \ldots \[then\] schools felt pressured to take them seriously" [@RN5007, p. 415]. Rankings *disciplined* the behavior of law schools. For example, applicants with characteristics valued by the rankings system (e.g., LSAT scores) became more important for decisions about admissions and merit aid [@RN5003]. Rankings also weakened the *authority* of admissions directors and admissions officers to make independent decisions about which applicants to admit, ideal class size, and curricular offerinags [@RN5003]. <!--
In a process akin to sports analytics in Moneyball \[CITE MICHAEL LEWIS\], we suggest that market research replaces the wisdom and authority of local admissions counselors.    -->

The analysis of UK school "league tables" by @RN5005 shows how quantification can be a mechanism of social reproduction. In 1992, the UK government began publishing school league tables, which ranked schools based on student performance on national exams taken at age 16. League tables facilitate making evaluative comparisons between schools regardless of geographic proximity. They also discipline households and schools to conceive of school quality in terms of test scores, which are substantially a function of the class composition of schools. Using Census data measured at the local authority level (similar to a U.S. county), @RN5005 find that localities with higher performing schools experienced growth in the share of managerial/professional households following the adoption of league tables and a decline in the share of working-class households. Using longitudinal survey data, managerial/professional households were more likely to move to localities with higher ranked schools following the introduction of school league tables. The introduction of league tables contributed to social reproduction because professional/managerial households were more aware of these consumer-facing metrics and had resources to respond by moving to more expensive neighbhorhoods, near higher performing schools.<!-- THIS IS RATCHET EFFECT STUFF. I CUT IT BECAUSE YOU HAVEN'T INTRODUCED THE IDEA OF RATCHET EFFECT
Schools that performed well on initial league tables tended to be schools with a high share of professional/managerial families. Therefore, quantification takes a snapshot of existing social stratification and that snapshot becomes a power unto itself, amplifying the effect of previous strctural inequality on subsequent structural inequality.
-->

```{=html}
<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].
-->
```
**Correlation**. The interdisciplinary literature on quantification [@RN5008] includes important contributions from the field of critical data studies [e.g., @RN4772; @RN4849]. In particular, the discussions of correlation (chapter 1) and homophily (chapter 2) by @RN4975 introduce ideas salient to the analysis the Market Segment Model and Geomarkets. Correlation measures the extent to which two or more variables move together. Predictive analytics are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared \ldots, big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'"[^2]

[^2]: Due to data limitations, predictive analytics often utilize "proxy" variables [@RN4849; @RN4975], which are variables that are correlated with a variable of interest. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of people like you. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access, when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?'

A bevy of studies show that predictions based on correlations reproduce structural inequality [for a review see @RN4794]. The correlations observed during the training data stage are a snapshot of relationships between variables at a particular point of time. The observed correlations may be a function of enduring structural inequality, but underlying causes are not considered by applications of predictive models. Reviewing scholarship about algorithms, @RN4794 [p. 224] state that "predicting the future on the basis of the past threatens to reify and reproduce existing inequalities." Disproportionately targeted/excluded populations are predicted to have a higher risk of an outcome, which amplifies subsequent targeting/exclusion. This phenomenon whereby has been termed the "ratchet effect" [@RN4851] and "pernicious feedback loops" [@RN4849].[^3] As we discuss below, @RN4982 analyzed 1980 score-sending behavior (training data) and inferred that demand for college is a function of class. They then recommended that selective colleges recruit from localities that have a high share of affluent, college-educated households.

[^3]: An often cited example is the LSI-R recidivism model which predicts a prisoner's chances of re-arrest and is used by 24 states [@RN4849]. Because the algorithm uses zip code as an input, people who live in highly policed neighborhoods have a higher predicted probability of being arrested, which leads to more policing in those neighborhoods, which perpetuates racialized inequality in arrests. Note that predictive models such as the LSI-R model are not merely used for social science purposes. Rather, they reproduce structural inequality because they direct the allocation of future resources based on a snapshot of historical inequality.

**Homophily**. Homophily is the idea actors who share common characteristics are likely to form connections with one another, or "birds of a feather flock together" [@mcpherson2001birds]. Homophily is a core concept of network science (writ large), in which actors (nodes) are connected to one another directly and indirectly via network ties (edges). As discussed below, @RN4982 [chapter 4] *The company We Keep: Colleges and Their Competition* conducts a network analysis based on 1980 SAT score sending behavior whereby two colleges are defined as competitors if a large number of students send SAT scores to both colleges. Because network science models often draws from rational choice theory, @RN4975 argues, it assumes that homophily is the result of voluntary action by individuals.

@RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. Like correlational models discussed above, because most network science models do not observe how historical structures affect relationships, they "erase historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that cause behavior consistent with homophily. Second, in commercial social networks, homopphily is more than an assumption; rather, it is programmed into algorithms that create connections between users. Thus, "social networks create and spawn the reality they imagine; they become self-fulfilling prophecies" [@RN4975, p. YY]. We observe similar processes in college access. Based on a snapshot of 1980 SAT score-sending, @RN4982 concludes that like-colleges compete for like-students (i.e., students of similar social origin). In turn, @RN4982 reasons that colleges should target the Geomarkets and high schools targeted by peer colleges. This logic is subsequently programmed into EPS software that recommends which Geomarkets and high schools to recruit from. Thus, homophily observed on the demand side -- which is a function of historic structural inequality –- is programmed into the supply-side.

Homophily is central to market research products that categorize customers. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].[^4] Classification situations engender markets where a vertical hierarchy of products are matched to a vertical hierarchy of consumers. We observe similar processes in our case study. @RN4982 categorizes students into four market segments -- local, in-state, regional, and national -- based on their score sending behavior and then evaluates the attractiveness of each Geomarket based on how many "regional" and "national" students live there. <!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

Clusters were given evocative names like "Blue Blood Estates," "\[Archie\] Bunker's Neighbors", and "Share Croppers." Because U.S. publicly available U.S. Census data disaggregated data only to the zip code-level, the Claritas Corporation had a strong incentive to argue that consumer behavior within zip codes was very similar [@RN4957]. @RN4565 emulated the behavior of Claritas by developing the *Segment Analysis Service: An Educationally Relevant Geodemographic Tagging Service*, which was an "add-on" to the *Student Search Service* student list product. Describing its conceptual underpinnings, @RN4565 [p. 1] states, "the basic tenet of geodemography is that people with similar cultural backgrounds, means, and perspectives naturally gravitate toward one another or form relatively homogeneous communities; in other words, birds of a feather flock together." The Segment product allocated each high school and each census tract into a distinct "educational neighborhoods" thought to reflect useful market segments for colleges looking for students.

PREVIOUS TEXT ABOUT CLASSIFICATION SITUATIONS
Related to the creation of market segments, @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Historically, classifications were binary. Consumers with "good" credit were offered loans and those with bad credit were not. The innovation of individual FICO scores -- and, later, other credit scores -- enabled merchants to classify consumers into many, fine-grained groups or along a continuum. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015]. People previously denied credit were now granted credit on different terms. For example, "payday loans" charge high interest rates to consumer groups that were previously denied credit altogether. Thus, at the bottom of the continuum, contemporary classification situations produce "predatory inclusion" -- exemplified by payday loans and for-profit colleges -- which is the logic of "including marginalized consumer-citizens into ostensibly democratizing mobility schemes on extractive terms" [@RN4774, p. 443].

Negative consequences of homophily
Chun (2021) discusses the negative consequences arising from homophily.Second, when we “engineer homophily” – through algorithms or policies – we make it less likely for heterophily to emerge, eliminating possibilities – of friendships, of matches – that may have occurred in the absence of such engineering.

Reviewing scholarship on quantification, @RN5006 argue that quantification has effects to the extent that stakeholders care about the numbers. U.S. law school rankings [@RN5003] and UK school league tables [@RN5005] exemplify this sort of salience. Studies of quantification in the sociology of education tend to study consumer-facing metrics. Law school rankings can be conceived as a business-to-consumer product because the goal is to inform consumers (prospective law school students) and consumers pay for the product.

-->

[^4]: For example, "payday loans" charge high interest rates to consumer groups that were previously denied credit altogether.

**Contribution**. As @RN3519 demonstrates in the chapter *Numbers*, enrollment management is fundamentally concerned with quantification. In the case of college rankings, sociology made visible how consumer-facing quantification disciplines the application/enrollment behavior of prospective students, the hiring behavior of employers, and the behavior of colleges who are disciplined to pursue customers with characteristics valued by rankings [@RN5003; @RN2356; @RN2375]. Beyond rankings, sociological analyses of enrollment management focus on the behavior of colleges and their agents, yielding the implicit assumption that inequalities creates by enrollment management are a function of individual college behavior. We argue that third-party vendors and products structure the enrollment management behavior of colleges. These products take a snapshot of student demand -- without considering structures that produce inequality -- and then recommend that colleges divert recruiting resources to localities with strong student demand. Based on a snapshot of existing social stratification, market research matches verticially categorized consumers to vertically categorized producers, thereby amplifying the effect of initial stratification on subsequent stratification. These products have been sold since the 1980s and have become more abundant with the ubiquity of software-as-service platforms, but have remained invisible to sociologists.

This manuscript analyzes Market Segment Model [@RN4982], which categorized high school students into vertical market segments and simultaneously created local Geomarkets that could be evaluated based on their composition of student market segments. The Market Segment Model and Geomarkets became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. Later, Geomarkets were incorporated into the College Board student list product named Student Search Service. Unlike the analysis of UK school league tables, @RN5005 we cannot show the effect of quantification. However, this study provides inisight about the mechanism underlying the effect. By analyzing and simulating *Student Search Service* purchases that filter on Geomarkets we show how Geomarkets reproduce historical race-based inequality in college access.

```{=html}
<!--
The market segment model became the basis for the College Board *Enrollment Planning Service* (EPS), which advised colleges which Geomarkets to target. @RN5021 reports that in 1995, 37% of 4-year public universities and 49% of 4-year private universities used EPS, while 41% of 4-year publics and 16% of 4-year privates used ACT's market analysis service product, which was similar to EPS. Geomarkets later became a filter in the College Board *Student Search Service* student list product, which is utilized by the vast majority of public and private universities.

By contrast, many quantification projects -- for example, human rights indicators -- have "little oomph" [@RN5006, p. 265]. FROM BERMAN & HIRSCHAN PAGE 263: "Merry demonstrates how human trafficking indicators largely sit on the shelf because no audience for them has been created." 
-->
```
# The Market Segment Model and College Board Geomarkets

```{=html}
<!--Historically, College Board partnered with Members to develop projects that later became revenue-generating products for College Board.

As a member organization, several College Board products developed from joint collaborations with administrators or professors from member institutions. During the 1970s, under the leadership of Larry Litten, then director of institutional research, Carleton College pionered applying market research to student recruiting. In the late 1970s, College Board provided Carleton College with funding and data for a six-year project that analyzed college choice decisions by households in six metropolitan areas. This project was the basis for the landmark book *Applying Market Research in College Admissions* [@RN4986], published by College Board, and was the basis for the College Board Admitted Student Questionnaire (ASQ), a revenue-generating product that provides colleges with information about admitted students -- those who enrolled and who did not enroll. Litten would later become Associate Director of the Consortium of Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s.

The Market Segment Model was an effort to quantify and communicate the knowledge that admissions officers have about student demand. 
-->
```
**Creating Geomarkets and the Market Segment Model.**

PUT THIS SOMEWHERE:

> Admissions officers invariably are tellers of stories -- about the colleges they represent, about the colleges they attended, about each other, and about the often vagabond life of college recruiting (p.9)\ldots We have begun with this celebration of sotrytelling\ldots \[because\] we believe that the intuitios of admissions officers actually comprise a remarkably systematic body of knowledge about the college selection process\ldots Our research this is based opn listening carefully to what admissions officers have to say (pp. 9-10).

In 1978, University of Pennsylvania professor Robert Zemsky, was tasked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x]. Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s. <!--The geographic concepts inscribed in the Market Segment Model were based on conversations with college admissions officers. --> The "initial task was to define enrollment markets in a manner consistent with admissions officers' intuitive understanding of student pools" [@RN4982,p. 4]. Quoting an admissions officer, @RN4982 [p. 11] write, "'There are only three kinds of college-bound students: those who want to live at home, those who want to live on campus but bring their laundry home, and those who want to go far enough from home that Mom and Dad can't visit without calling first.'" As such, [@RN4982] created three geographic [@RN4982, p. 11], region, state, and community. The three regions were New England, Middle States, and the South. They "divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets" (p. 11). These enrollment markets, later called Geomarkets, were intended to be consistent with the conception of a catchment market from the perspective of admissions counselors. @RN4982 [pp. 11-12] briefly describes the creation of Geomarket borders:

> In many cases, the market boundaries match formal political and educational divisions, reflecting natural channels of communication. Each major metropolitan area is composed of several markets, usually corresponding to the inner city, a first ring of suburbs, and an outer ring of suburbs. In more sparsely populated areas, communities are sometimes combined in order to make the analysis meaningful."

>>>>

```{=tex}
\newpage
\pagenumbering{gobble}
\pagenumbering{arabic}
```
# Data and Methods
